---
title: "Getting Started with conformeR"
author: "Justine Leclerc and Christof Seiler"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with conformeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  fig.width = 12,
  fig.height = 7,
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)
```

```{css, echo = FALSE}
.callout {
  border-left: 4px solid #4472A8;
  background-color: #F2F6FA;
  padding: 0.6em 1em;
  margin: 1em 0;
  border-radius: 2px;
}
.callout p {
  margin: 0.3em 0;
}
```

## Overview

`LEMUR` fits a low-dimensional embedding and assigns each cell to a
*neighborhood* when they share a similar response to a given contrast and gene. Because this
assignment is itself the output of a fitted model, it inherits sampling
variability: refit the model on a slightly different draw of cells, and
the neighborhood boundaries shift. `conformeR` wraps this neighborhood
construction step in a conformal prediction layer so that this
variability can be quantified rather than ignored.

It provides two complementary views of that uncertainty:

- **Conformal selection** asks, for a given gene, *which cells can we
  confidently say belong to the affected neighborhood?* It returns a
  data-driven subset of cells with a guaranteed bound on the
  proportion of false inclusions.
- **Conformal clustering** asks the finer-grained question, *for each
  cell, how uncertain are we about its neighborhood assignment?* It
  returns a prediction *set* of plausible neighborhoods per cell,
  which can contain zero, one, or several labels.

<div class="callout">
<p>**Why this matters.** A single LEMUR fit gives you one particular
neighborhood assignment, with no way to tell how much of it would
survive resampling. `conformeR` turns that single assignment into a
statement with a formal, finite-sample coverage guarantee.</p>
</div>

### A caveat: the guarantee assumes true labels

The coverage guarantees behind conformal prediction are proven under
the assumption that the training and calibration sets carry the
*true* neighborhood label for each cell. That assumption is exactly
what motivates reaching for conformal prediction here in the first
place — but it is also never satisfied in practice: we do not observe
the true neighborhood membership of any cell. What we have instead is
LEMUR's own predicted neighborhood assignment, which we use as a
proxy label for both training and calibration. That proxy is a
*noisy* version of the true label, and the two conformal procedures
degrade differently under that noise:

- **Conformal clustering** stays conservative. Calibrating against a
  noisy label inflates the non-conformity scores, so the procedure
  compensates by returning more prediction sets of size 2 — i.e. it
  more often declines to commit to a neighborhood — rather than
  silently under-covering. The formal guarantee effectively still
  holds, at the cost of more "I don't know" answers.
- **Conformal selection** has no such built-in safety valve. Its
  validity relies directly on LEMUR's predicted neighborhoods not
  being *too* wrong: if LEMUR's assignment is systematically biased
  rather than just noisy, the selected neighborhood inherits that
  bias. In practice this means conformal selection results are only
  as trustworthy as the underlying LEMUR fit.

<div class="callout">
<p>**Practical takeaway.** Read a size-2 prediction set from conformal
clustering as the procedure protecting itself against label noise,
not as evidence that a cell has no meaningful neighborhood. Read a
conformal selection result with an eye on LEMUR fit quality first —
e.g. convergence diagnostics, embedding dimensionality — since that
fit stands in for ground truth throughout.</p>
</div>

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
if (!requireNamespace("conformeR", quietly = TRUE)) {
  devtools::install_github("https://github.com/juslecl/conformeR")
}
library(conformeR)
library(dplyr)
library(ggplot2)
library(tibble)
library(knitr)
```

We use a consistent, minimal theme throughout the vignette so that the
`conformeR` plots line up visually with the tables:

```{r theme}
theme_set(
  theme_minimal(base_size = 13) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold"),
      legend.position = "bottom"
    )
)
conformer_palette <- c("inside" = "#4472A8", "outside" = "#D9822B", "none" = "#B0B0B0")
```

## Example dataset

This vignette runs `conformeR` on a reduced version of the single-cell
RNA-seq dataset from Zhao, Dovas, Spinazzi, et al. (2021),
*"Deconvolution of cell type-specific drug responses in human tumor
tissue with single-cell RNA-seq,"* published in *Genome Medicine*
(GEO accession `GSE148842`). As with any input to `conformeR`, the
data have already been preprocessed and quality-checked upstream.

```{r load-data}
data("sce_vignette")
sce_vignette
head(colData(sce_vignette))
```

## Selecting genes of interest

We restrict the analysis to three genes with distinct expected
behavior, to illustrate a range of neighborhood sizes and
uncertainty patterns: `HIST3H2A`, `PTPRZ1`, and `HBEGF`.

```{r genes-of-interest}
pan_affect_genes <- rowData(sce_vignette) %>%
  as_tibble() %>%
  dplyr::select(gid, gene) %>%
  filter(gene %in% c("HIST3H2A", "PTPRZ1", "HBEGF"))

kable(pan_affect_genes, caption = "Genes of interest for this vignette")
```

## Running conformeR

`conformeR()` fits `LEMUR` and layers both conformal procedures on
top of it in one call. The key arguments are:

- `design_lemur` — the LEMUR design formula; here we adjust for
  `patient_id` and estimate the effect of `condition`.
- `contrast_column` — which column of the design defines the
  contrast of interest.
- `genes_of_interest` — restrict the (more expensive) conformal
  layers to a gene subset; LEMUR itself is still fit genome-wide.
- `what` — which conformal procedure(s) to run: `"conf_selection"`,
  `"conf_clustering"`, or both.

By default the target coverage is `cp_alpha`, and the data are split
50% training / 50% calibration+test, with the calibration set taking
25 percentage points of the total (`cp_train_frac` and `cp_cal_frac`
override these proportions).

```{r run-conformer}
set.seed(2002)
res <- conformeR(
  sce_vignette,
  design_lemur = ~ patient_id + condition,
  contrast_column = "condition",
  genes_of_interest = pan_affect_genes$gid,
  what = c("conf_selection", "conf_clustering"),
  n_cores = 1,
  verbose = TRUE
)
```

```{r tidy-results}
cp_res <- res$conf_results |>
  left_join(pan_affect_genes, by = c("gene" = "gid")) |>
  dplyr::select(-gene) |>
  dplyr::rename(gene = gene.y)
```

## Conformal selection: how big is the affected neighborhood?

For each gene, conformal selection returns the set of cells flagged
as belonging to the neighborhood affected by `condition`, at the
target coverage level. The table below reports both the raw
neighborhood size and its share of all cells in the fitted embedding.

```{r selection-summary}
selection_summary <- cp_res |>
  group_by(gene) |>
  summarize(
    `Neighborhood size` = sum(inside_cs),
    `% of cells` = round(sum(inside_cs) / ncol(res$fit_lemur), 2) * 100,
    .groups = "drop"
  )

kable(selection_summary, caption = "Conformal selection: neighborhood size per gene")
```

A larger `% of cells` means the gene's condition effect is detected
across a broader swath of the embedding; a small, tight neighborhood
points to a more localized effect.

## Conformal clustering: how confident is each cell's assignment?

Conformal clustering instead reports, per cell, a *prediction set*:
the neighborhoods that cell could plausibly belong to at the target
coverage. A prediction set of size zero means the procedure could not
confidently assign that cell to any neighborhood; larger sets reflect
more ambiguous cells.

```{r clustering-summary}
clustering_summary <- cp_res |>
  group_by(gene) |>
  summarize(
    `Prediction set = inside`  = sum(inside_cc == 1),
    `Prediction set = outside` = sum(outside_cc == 1),
    `Prediction set size 2`    = sum(inside_cc + outside_cc == 0),
    .groups = "drop"
  )

kable(clustering_summary, caption = "Conformal clustering: prediction set composition per gene")
```

Reading this table alongside the selection table above shows where
the two procedures agree (a gene with a large, confident
neighborhood in both) and where they diverge (a gene with a sizeable
selected neighborhood but many size-2 or ambiguous cells under
clustering, which flags instability worth a closer look). Recall from
the caveat above that a high count of size-2 sets is, by design, the
clustering procedure's response to calibrating against LEMUR's noisy
neighborhood labels rather than the true ones — so treat it as a
built-in conservatism, not necessarily a sign that something is wrong
with the gene's effect.

## Visualizing results on the LEMUR embedding

### Conformal selection

```{r plot-selection, fig.cap = "Conformal selection overlaid on LEMUR neighborhoods, one panel per gene."}
plotter_conformal_selection(res, pan_affect_genes$gid)
```

Cells shaded as part of the selected neighborhood are those the
procedure can confidently attribute to the condition effect for that
gene; unshaded cells fall outside the guaranteed selection at the
chosen coverage level.

### Conformal clustering

```{r plot-clustering, fig.cap = "Conformal prediction sets overlaid on LEMUR neighborhoods, one panel per gene."}
plotter_conformal_clustering(res, pan_affect_genes$gid)
```

Here color reflects the prediction set itself rather than a single
hard assignment, making it easy to spot the boundary regions where
neighborhood membership is genuinely uncertain versus the core
regions where every resampling agrees.

## Session info

```{r session-info}
sessionInfo()
```
